---
title: "DATA 606 Chapter 3 - Probability"
author: 'David Simbandumwe'
output:
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("VennDiagram")

```

**Dice rolls.** (3.6, p. 92) If you roll a pair of fair dice, what is the probability of


```{r }

#b)
4/36

#c)
1/36

```



(a) getting a sum of 1?

**0% probability 2 is the lowest sum you can get with a pair of dice**

(b) getting a sum of 5?

**4/36 or 11.1%**

(c) getting a sum of 12?

**1/36 or 2.8%**



--------------------------------------------------------------------------------

\clearpage

**Poverty and language**. (3.8, p. 93) The American Community Survey is an ongoing survey that provides data every year to give communities the current information they need to plan investments and services. The 2010 American Community Survey estimates that 14.6% of Americans live below the poverty line, 20.7% speak a language other than English (foreign language) at home, and 4.2% fall into both categories.

(a) Are living below the poverty line and speaking a foreign language at home disjoint?

**no 4.2% fall into both categories**

(b) Draw a Venn diagram summarizing the variables and their associated probabilities.

```{r venn diagram}

grid.newpage()
  
draw.pairwise.venn(area1=14.6, area2=20.7,cross.area=4.2,
                   category=c("below poverty","foreign language"),fill=c("Red","Yellow"))



```



(c) What percent of Americans live below the poverty line and only speak English at home?

**10.4%**


(d) What percent of Americans live below the poverty line or speak a foreign language at home?

**4.2%**

(e) What percent of Americans live above the poverty line and only speak English at home?

**68.9%**

(f) Is the event that someone lives below the poverty line independent of the event that the person speaks a foreign language at home?

**there is a higher likelihood of poverty for foreign language speakers 25.5% to 15.1%**




--------------------------------------------------------------------------------

\clearpage

**Assortative mating**. (3.18, p. 111) Assortative mating is a nonrandom mating pattern where individuals with similar genotypes and/or phenotypes mate with one another more frequently than what would be expected under a random mating pattern. Researchers studying this topic collected data on eye colors of 204 Scandinavian men and their female partners. The table below summarizes the results. For simplicity, we only include heterosexual relationships in this exercise.

\begin{center}
\begin{tabular}{ll  ccc c}
                                        &           & \multicolumn{3}{c}{\textit{Partner (female)}} \\
\cline{3-5}
                                        &           & Blue  & Brown     & Green     & Total \\
\cline{2-6}
                                        & Blue      & 78    & 23        & 13        & 114 \\
\multirow{2}{*}{\textit{Self (male)}}   & Brown     & 19    & 23        & 12        & 54 \\
                                        & Green     & 11    & 9         & 16        & 36 \\
\cline{2-6}
                                        & Total     & 108   & 55        & 41        & 204
\end{tabular}
\end{center}


(a) What is the probability that a randomly chosen male respondent or his partner has blue eyes?

**70.6%**

(b) What is the probability that a randomly chosen male respondent with blue eyes has a partner with blue eyes?

**38.2%**

(c) What is the probability that a randomly chosen male respondent with brown eyes has a partner with blue eyes? 

**9.3%**

What about the probability of a randomly chosen male respondent with green eyes having a partner with blue eyes?

**5.4%**

(d) Does it appear that the eye colors of male respondents and their partners are independent? Explain your reasoning.

**It appears that if a male respondent has blue eyes they are more likely to have a partner with blue eyes 38.2% vs 9.3% and 5.4% for male partners with brown and green eyes respectivily.** 




--------------------------------------------------------------------------------

\clearpage

**Books on a bookshelf**. (3.26, p. 114) The table below shows the distribution of books on a bookcase
based on whether they are nonfiction or fiction and hardcover or paperback.

\begin{center}
\begin{tabular}{ll  cc c}
                                &           & \multicolumn{2}{c}{\textit{Format}} \\
\cline{3-4}
                                &           & Hardcover     & Paperback     & Total \\
\cline{2-5}
\multirow{2}{*}{\textit{Type}}  & Fiction   & 13            & 59            & 72 \\
                                & Nonfiction& 15            & 8             & 23 \\
\cline{2-5}
                                & Total     & 28            & 67            & 95 \\
\cline{2-5}
\end{tabular}
\end{center}

(a) Find the probability of drawing a hardcover book first then a paperback fiction book second when drawing without replacement.

**18.3%**

(b) Determine the probability of drawing a fiction book first and then a hardcover book second, when drawing without replacement.

**21.8%**

(c) Calculate the probability of the scenario in part (b), except this time complete the calculations under the scenario where the first book is placed back on the bookcase before randomly drawing the second book.

**22.3%**

(d) The final answers to parts (b) and (c) are very similar. Explain why this is the case.

**without replacement - if you select a fiction book in the first drawing then there is one less book available when attempt to select a hard cover book.**




--------------------------------------------------------------------------------

\clearpage

**Baggage fees**. (3.34, p. 124) An airline charges the following baggage fees: $25 for the first bag and $35 for the second. Suppose 54% of passengers have no checked luggage, 34% have one piece of checked luggage and 12% have two pieces. We suppose a negligible portion of people check more than two bags.

(a) Build a probability model, compute the average revenue per passenger, and compute the corresponding standard deviation.
+ average revenue = 12.7
+ std deviation = 15.85

(b) About how much revenue should the airline expect for a flight of 120 passengers? With what standard deviation? Note any assumptions you make and if you think they are justified.

**expected revenue for 120 passangers is $1,524 with a std deviation of $1,902. Assuming that passangers on the flight are a represenation of the population in general.**




--------------------------------------------------------------------------------

\clearpage

**Income and gender**. (3.38, p. 128) The relative frequency table below displays the distribution of annual total personal income (in 2009 inflation-adjusted dollars) for a representative sample of 96,420,486 Americans. These data come from the American Community Survey for 2005-2009. This sample is comprised of 59% males and 41% females.

\begin{center}
\begin{tabular}{lr}
  \hline
\textit{Income}         & \textit{Total} \\
  \hline
\$1 to \$9,999 or loss  & 2.2\% \\
\$10,000 to \$14,999    & 4.7\% \\
\$15,000 to \$24,999    & 15.8\% \\
\$25,000 to \$34,999    & 18.3\% \\
\$35,000 to \$49,999    & 21.2\% \\
\$50,000 to \$64,999    & 13.9\% \\
\$65,000 to \$74,999    & 5.8\% \\
\$75,000 to \$99,999    & 8.4\% \\
\$100,000 or more       & 9.7\% \\
   \hline
\end{tabular}
\end{center}

(a) Describe the distribution of total personal income.

**normally distributed**

(b) What is the probability that a randomly chosen US resident makes less than $50,000 per year?

**62.2%**

(c) What is the probability that a randomly chosen US resident makes less than $50,000 per year and is female?
Note any assumptions you make.

**25.5% if we assume that the income distribution of the female population is that same as the income distribution for the general population **

(d) The same data source indicates that 71.8% of females make less than $50,000 per year. Use this value to determine whether or not the assumption you made in part (c) is valid.

**given that 62.2% of the general population makes less than $50,000 it appears that the distribution of income for the female population is not consistent with the general population.**




