---
title: "DATA 606 Data Project: Causes of Financial Wellbeing"
author: "David Simbandumwe"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
editor_options: 
  chunk_output_type: inline
---


```{r, echo=FALSE, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(utils)

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim)
library(dplyr)
library(plyr)
library(naivebayes)

library(glue)
library(GGally)
library(kableExtra)
library(vip)

```



## Setup

```{r clean, results="hide", include=FALSE}

set.seed(1234)
rm(list=ls())

```



```{r external_functions, results="hide",  error=TRUE}

# Exported data load, tidy function and tibble cleaning
fun_file_name <- glue(getwd(), "/Project/DATA606_Functions.R")
source(fun_file_name, local = knitr::knit_global())

```





# Abstract
The deregulation of the financial services industry has provided more investment opportunities for individuals and families but it has also increased the level of financial knowledge necessary to make good decisions. It is no coincidence that these trends have driven an increased focus in financial health and financial wellbeing by individuals, employers, government agencies and financial services firms. Understanding financial wellness requires a clear definition of wellness, an approach to measuring differences in wellbeing, and an understanding the factors that influence consumer financial wellness.

This project adopts the definition and measurement methodology developed by the Consumer Financial Protection Bureau as part of their ongoing research efforts. The CFPB defines financial wellness as the “...sense of financial security and freedom of choice—both in the present and when considering the future”. The CFPB has developed a measurement of consumer financial wellbeing called the Financial Wellbeing Scale (CFPB Score). The CFPB methodology includes a survey that not only captures the 10 questions necessary to calculate the CFPB score but includes additional attributes that paint a holistic financial and demographic picture of an individual or family. 

This analysis will utilize the data from the CFPB Survey. Starting with the non calculated variables linear regression will be used to evaluate the relationship between non scoring attributes demographic, financial, house hold structure characteristics and the CFPB Score. 

(Please note that the 10 questions that are used to calculate the CFPB Score are not included in the attributes evaluated in this analysis)


# Part 1 - Introduction

This analysis is focused on understanding the factors that impact consumer financial wellbeing. There are numerous sources of analysis, theory and survey data however for the purposes of this project I will focus on the work from the Consumer Financial Protection Bureau National Financial Well-Being Survey Public Use File (PUF). The PUF survey data allows us to analyze a CFPB Score of consumer wellbeing in association with an individuals financial and demographic characteristics. Linear regression will be used to quantify the impact of individual factors and create a predictive model of financial wellbeing. 

<br>
**Analysis Steps**
In addition to the outline provided for the project that analysis will follow that steps below.

1. Review the survey data including the cfpb score 
2. Select subset of 217 variables in the survey to analyze 
3. Explore the relationship between individual variables and the cfpb score
4. Fit the full model using linear regression
5. Optimize the model using stepwize regression
6. Fit a models focused on financial planning variables and demographic variables respectively
7. Explore impacts of survey variables on cfpb score



# Part 2 - Data
The data was collected as part of the Consumer Financial Protection Bureau’s (CFPB) National Financial Well-Being Survey Public Use File (PUF). The PUF is a dataset containing 

(1) data collected in the National Financial Well-Being Survey, 
(2) data about members of the GfK KnowledgePanel collected prior to the survey, and 
(3) data on poverty levels in respondents’ counties of residence.

The National Financial Well-Being Survey was conducted in English and Spanish via web mode between October 27, 2016 and December 5, 2016. Overall, 6,394 surveys were completed: 5,395 from the general population sample and 999 from an oversample of adults aged 62 and older. The survey was designed to represent the adult population of the 50 U.S. states and the District of Columbia. The survey was fielded on the GfK KnowledgePanel®. The KnowledgePanel sample is recruited using address-based sampling and dual-frame landline and cell phone random digit dialing methods.

The PUF was published in 2017.




**Load data and review summary stats**

Limited filtering of observations conducted as part of this analysis  Other answers such as "refuse to answer" or "do not know" actual contain some relevant information in the context of a individuals financial characteristics. This can be seen in the prevalence of these answers in the dataset roughly 2/3 or the observations include at least one survey answer in the other category.   

```{r}

# full dataset
cfpbRaw_df <- getRawCFPBFile()


# load data - logic for data tidying is associated DATA606_Functions.R file
cfpb_df <- getReducedCFPBFile()
cfpb_df <- cfpb_df %>% droplevels()


# Filter records with invalid cfpb scores
cfpb_df <- cfpb_df %>% filter( cfpb_score > 0)
    
```


```{r}

summary(cfpb_df)

```

The cfpb score for this survey data is near normaly distributed with outlines at the high end and low end of the scoring distribution. 

```{r}

# cfpb
ggplot(cfpb_df,
    aes(cfpb_score, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", color="red", size = 0.5) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (cfpb dataset)")
summary(cfpb_df$cfpb_score)

```

<br>
**House Hold Income**

As would be expected the histogram and the box plot for cfpb score categorized by house hold income shows some positive correlation between house hold income and the cfpb_score. 

```{r}


# econ_hh_income 
ggplot(cfpb_df, aes(x=econ_hh_income)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "econ_hh_income")

ggplot(data = cfpb_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=econ_hh_income, color=econ_hh_income)) +
  labs (title="CFPB Score ( by econ_hh_income)", x="CFPB score", y="house hold income")


ggplot(cfpb_df,
    aes(cfpb_score, fill = econ_hh_income, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by econ_hh_income)")

```


<br>
**Age**

As would be expected the histogram and the box plot for cfpb score categorized by age shows some positive correlation between house hold income and the cfpb_score. This finding aligns with the typical relationship between networth, income and age. 

```{r}


# dem_age 
ggplot(cfpb_df, aes(x=dem_age)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "dem_age")

ggplot(data = cfpb_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=dem_age, color=dem_age)) +
  labs (title = "CFPB Score (by dem_age)", x="CFPB score", y="age category")


ggplot(cfpb_df,
    aes(cfpb_score, fill = dem_age, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by dem_age)")


```

<br>
**Achieve Financial Goals**

The relationship between sfpb_score and confidence in reaching your financial goals is intriguing It highlights a positive relationship between confidence in financial goals and cfpb score. The box plot and histogram show a wide range of survey respondents are optimistic even if their cfpb scores would indicate otherwise.  


```{r}

# fin_achieve_goals
ggplot(cfpb_df, aes(x=fin_achieve_goals)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "fin_achieve_goals")

ggplot(data = cfpb_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=fin_achieve_goals, color=fin_achieve_goals)) +
  labs (title = "CFPB Score (by fin_achieve_goals)", x="CFPB score", y="belief in achieving fin goals")


ggplot(cfpb_df,
    aes(cfpb_score, fill = fin_achieve_goals, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by fin_achieve_goals)")

```


<br>
**Health**

Given the structure of the health care system in the United States it is unsurprising that health shows a positive correlated cfpb score. 


```{r}

# per_health
ggplot(cfpb_df, aes(x=fin_frugal)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "per_health")

ggplot(data = cfpb_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=per_health, color=per_health)) +
  labs (title = "CFPB Score (by per_health)", x="CFPB score", y="personal health evaluation")


ggplot(cfpb_df,
    aes(cfpb_score, fill = per_health, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by per_health)")


```





# Part 3 - Exploratory data analysis

This analysis uses linear regression to identify and quantify the impacts of survey variables to the cfpb score. 

<br>
**Preconditions for Linear Models**

The residual analysis from the CFPB model and Federal Reserve model roughly adheres to the following pre-conditions for linear modeling. More analysis would be required to ensure independence and a linear relationship to the dependent variable. It would be difficult to assert full independence of the predictor variables given that each case reflect an individual. Thre are numerous studies that link demographics, race, geography and age to economic outcomes. 

1. residuals are nearly normal
2. residuals have constant variability
3. residuals are independent
4. each variable is linearly related to the outcome





## Fit Full Model

The full model of all 26 variables included in the analysis is used as a starting point. The resulting model has an adjusted $R^2$ of 0.5739.

```{r}

# split data
cfpb_split <- initial_split(cfpb_df, prop = 0.8, strata = cfpb_score)
cfpb_training <-  training(cfpb_split)
cfpb_test <-  testing(cfpb_split)



# fit full model
cfpb_result1<-lm(cfpb_score ~ . ,data=cfpb_training)
summary(cfpb_result1)
 
 
```





## Stepwise model optimization

Starting with the full model that incldued all 26 variables setpwize regression was used to optimize the model. 

```{r, echo=FALSE, results='hide', include=FALSE}

# Bayes information criterion (BIC)
cfpb_result2<-stats::step(cfpb_result1, k=log(nrow(cfpb_training)))

```






# Part 4 - Inference


## Optimized Model


Model stats
- The adjuste $R^2$ for the model is 0.5633
- The residual analysis analysis shows several outliers that are impacting the model


```{r}

# Step:  AIC=23075.37
# cfpb_score ~ econ_hh_income + econ_savings + per_health + dem_age + 
#     fin_achieve_goals + fin_saving_habit + fin_consult_budget + 
#     fin_set_goals


# Model Specification
lm_model <- linear_reg() %>% 
            set_engine('lm') %>% 
            set_mode('regression')


# Fitting to Trained Data
lm_fit <- lm_model %>% 
          fit(cfpb_score ~ econ_hh_income + econ_savings + per_health + dem_age + 
                            fin_achieve_goals + fin_saving_habit + fin_consult_budget + 
                            fin_set_goals
              
              , data = cfpb_training)


# Explore Training Results
names(lm_fit)
summary(lm_fit$fit)
par(mfrow=c(2,2)) 
plot(lm_fit$fit, pch = 16, col = '#006EA1')


# Data frame of estimated coefficients
tidy(lm_fit)

# Performance metrics on training data
glance(lm_fit)  

# variable importance
vip(lm_fit, num_features=15)

```



<br>
**Testing Optimized Model**


```{r}

#predict(lm_fit, new_data = cfpb_test)

cfpb_results <- predict(lm_fit, new_data = cfpb_test) %>% 
                            bind_cols(cfpb_test)

#cfpb_results

# RMSE on test set
rmse(cfpb_results, truth = cfpb_score, estimate = .pred)
rsq(cfpb_results, truth = cfpb_score, estimate = .pred)



```




```{r}

ggplot(data = cfpb_results,
       mapping = aes(x = .pred, y = cfpb_score)) +
  geom_point(color = '#006EA1') +
  geom_smooth(method = lm) +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +
  labs(title = 'Linear Regression Results - Optimized',
       x = 'Predicted',
       y = 'Actual')

```




```{r}

# econ_hh_income
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=econ_hh_income, color=econ_hh_income)) +
  labs (title = "econ_hh_income")


# econ_savings
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=econ_savings, color=econ_savings)) +
  labs (title = "econ_savings")

# per_health
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=per_health, color=per_health)) +
  labs (title = "per_health")


# dem_age
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=dem_age, color=dem_age)) +
  labs (title = "dem_age")


# fin_achieve_goals
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=fin_achieve_goals, color=fin_achieve_goals)) +
  labs (title = "fin_achieve_goals")


# fin_saving_habit
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=fin_saving_habit, color=fin_saving_habit)) +
  labs (title = "fin_saving_habit")

# fin_consult_budget
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=fin_consult_budget, color=fin_consult_budget)) +
  labs (title = "fin_consult_budget")

# fin_set_goals
ggplot(data = cfpb_results) +
  geom_point(mapping = aes(x=cfpb_score, y=fin_set_goals, color=fin_set_goals)) +
  labs (title = "fin_set_goals")


```




```{r}

# cfpb

cfpb_results$.residual <- cfpb_results$cfpb_score - cfpb_results$.pred


ggplot(cfpb_results,
    aes(.residual, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(.residual)), linetype = "dashed", color="red", size = 0.5) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Residual Histogram")
summary(cfpb_df$cfpb_score)



cfpb_results %>%
#    filter(cfpb_score > 47 & cfpb_score < 66) %>%
    ggplot(
        aes(y = abs(.residual), x = .pred)) +
        geom_point(alpha = 0.7, color="blue",size = 0.1) +
        geom_smooth(method = lm) + 
        #geom_vline(aes(xintercept = mean(.residual)), linetype = "dashed", color="red", size = 0.5) +
        scale_fill_brewer(palette="Spectral") +
        labs(title = "CFPB Residual (vs predictions)")
summary(cfpb_df$cfpb_score)




```


```{r}

# econ_hh_income
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=econ_hh_income, color=econ_hh_income)) +
  labs (title = "econ_hh_income")


# econ_savings
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=econ_savings, color=econ_savings)) +
  labs (title = "econ_savings")

# per_health
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=per_health, color=per_health)) +
  labs (title = "per_health")


# dem_age
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=dem_age, color=dem_age)) +
  labs (title = "dem_age")


# fin_achieve_goals
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=fin_achieve_goals, color=fin_achieve_goals)) +
  labs (title = "fin_achieve_goals")


# fin_saving_habit
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=fin_saving_habit, color=fin_saving_habit)) +
  labs (title = "fin_saving_habit")

# fin_consult_budget
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=fin_consult_budget, color=fin_consult_budget)) +
  labs (title = "fin_consult_budget")

# fin_set_goals
ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.residual, y=fin_set_goals, color=fin_set_goals)) +
  labs (title = "fin_set_goals")


```






##  Demographic Model

The demographic model that includes race, gender and education has limited predictive power for overall cfpb score on its own It would be interesting to review the impacts of race, gender and education if we held other variables constant. 

Model stats
- r-squared for the model is 0.08288
- The residual analysis analysis shows several outliers that are impacting the model



```{r}

# Step:  AIC=22814.15
# cfpb_score ~  dem_edu + dem_race +dem_gender   


# Model Specification
lm_model <- linear_reg() %>% 
            set_engine('lm') %>% 
            set_mode('regression')


# Fitting to Trained Data
lm_fit <- lm_model %>% 
          fit(cfpb_score ~ dem_edu + dem_race + dem_gender 
              
              , data = cfpb_training)


# Explore Training Results
names(lm_fit)
summary(lm_fit$fit)
par(mfrow=c(2,2)) 
plot(lm_fit$fit, pch = 16, col = '#006EA1')


# Data frame of estimated coefficients
tidy(lm_fit)

# Performance metrics on training data
glance(lm_fit)  

# variable importance
vip(lm_fit, num_features=15)

```

<br>
**Testing Demographic Model**

```{r}

#predict(lm_fit, new_data = cfpb_test)

cfpb_results <- predict(lm_fit, new_data = cfpb_test) %>% 
                            bind_cols(cfpb_test)

#cfpb_results

# RMSE on test set
rmse(cfpb_results, truth = cfpb_score, estimate = .pred)
rsq(cfpb_results, truth = cfpb_score, estimate = .pred)


```

    

```{r}

ggplot(data = cfpb_results,
       mapping = aes(x = .pred, y = cfpb_score)) +
  geom_point(color = '#006EA1') +
  geom_smooth(method = lm) +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +
  labs(title = 'Linear Regression Results - Demographic',
       x = 'Predicted',
       y = 'Actual')

```
    

## Financial Planning

The model that focused on financial planning, budgeting and setting goals has some explanatory power. Although 0.2717 is relativily low in comparison to the other models it is significant when considered in the context of the variable that are not included such as income, education and age. 

Model stats
- r-squared for the model is 0.2717
- The residual analysis analysis shows several outliers that are impacting the model


```{r}

# Step:  AIC=22814.15
# cfpb_score ~  fin_goals + fin_saving_habit + fin_frugal + 
#                       fin_consult_budget + fin_follow_budget + fin_set_goals + fin_clear_plan   


# Model Specification
lm_model <- linear_reg() %>% 
            set_engine('lm') %>% 
            set_mode('regression')


# Fitting to Trained Data
lm_fit <- lm_model %>% 
          fit(cfpb_score ~ fin_goals + fin_saving_habit + fin_frugal + fin_consult_budget + 
                  fin_follow_budget + fin_set_goals + fin_clear_plan  
              
              , data = cfpb_training)


# Explore Training Results
names(lm_fit)
summary(lm_fit$fit)
par(mfrow=c(2,2)) 
plot(lm_fit$fit, pch = 16, col = '#006EA1')


# Data frame of estimated coefficients
tidy(lm_fit)

# Performance metrics on training data
glance(lm_fit)  

# variable importance
vip(lm_fit, num_features=15)

```



<br>
**Testing Financial Planning Model**



```{r}

#predict(lm_fit, new_data = cfpb_test)

cfpb_results <- predict(lm_fit, new_data = cfpb_test) %>% 
                            bind_cols(cfpb_test)

#cfpb_results

# RMSE on test set
rmse(cfpb_results, truth = cfpb_score, estimate = .pred)
rsq(cfpb_results, truth = cfpb_score, estimate = .pred)


```

    

```{r}

ggplot(data = cfpb_results,
       mapping = aes(x = .pred, y = cfpb_score)) +
  geom_point(color = '#006EA1') +
  geom_smooth(method = lm) +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +
  labs(title = 'Linear Regression Results - CFPB Score',
       x = 'Predicted',
       y = 'Actual')

```
    

```{r}


# age category

cfpb_results$cfpb_score_10cat <- cut(cfpb_results$cfpb_score, breaks = c(-10, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
                           labels = c("< 10","10-20","20-30","30-40","40-50","50-60","60-70","70-80","80-90","90-100"),
                           right = FALSE,
                           include.lowest=TRUE) 

t <- cfpb_results %>% select(cfpb_score,cfpb_score_10cat)

ggplot(cfpb_results,
    aes(.pred, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(.pred)), linetype = "dashed", color="red", size = 0.5) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (.pred)")
summary(cfpb_df$.pred)


ggplot(data = cfpb_results) +
  geom_boxplot(mapping = aes(x=.pred, y=cfpb_score_10cat, color=cfpb_score_10cat)) +
  labs (title = "cfpb_score_10cat")


ggplot(cfpb_results,
    aes(.pred, fill = cfpb_score_10cat, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(.pred)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by cfpb_score_10cat)")

```



# Part 5 - Conclusion

A linear regression model was developed using the results of the Consumer Financial Protection Bureau’s (CFPB) National Financial Well-Being Survey Public Use File (PUF). The model predicts over 0.5608 cfpb score given the 8 variable selected for the optimized model. 

**Model Performance**
- adjusted $R^2$ 0.5608655	
- root mean square error 9.609951

**Model Variables**
The following variable are included in the model as significant contributors to the overall model accuracy. 

- econ_hh_income (PPINCIMP) - The survey respondents Household Income for a given year.
- econ_savings (SAVINGSRANGES) - An assessment of the amount of savings that the survey respondant currenlty has. 
- per_health (HEALTH) - An assessment of the survey respondents overall health.
- dem_age (agecat) - The age of the survey respondent (8 categories).
- fin_achieve_goals (GOALCONF) - An assessment of the ability of the survey respondent to achieve their financial goals. 
- fin_saving_habit (SAVEHABIT) - An assessment of savings habits and how often money is put away.
- fin_consult_budget (PROPPLAN_1) - An assessment of how budgeting impacts spending decisions.
- fin_set_goals (PROPPLAN_3) - An assessment of financial planning and goal setting discipline.


The findings for the analysis are not surprising and are consistent with several studies in the industry however deeper analysis is warranted. The cfpb scaled score is designed to cover the entire population however it is not surprising that the resulting model breaks down slightly at the higher and lower ranges of the cfpb score. A more localized analysis of cfpb score ranges might provide some additional insight into the factors that impact consumer financial wellness.  



# References

There are numerous studies of financial wellness sponsored by financial firms, industry groups and government agencies. I choose to focus on two studies:

- Consumer Financial Protection Bureau - Financial Wellness Survey [*Financial well-being data*](https://www.consumerfinance.gov/data-research/financial-well-being-survey-data/) The PUF survey results can be accessed as a csv file [*Financial well-being survey data*](https://www.consumerfinance.gov/documents/5614/NFWBS_PUF_2016_data.csv)


# Appendix (optional)


## CFPB Financial Well-Being Scale Methodology

The CFPB Financial Well-Being Scale methodology uses the following 10 questions to calculate the cfpb score. There is some overlap between the questions used to calculate the cfpb score and the survey questions. The optimized models presented in this analysis did not include any of the variable that overlap with the scoring questions. 

How well does this statement describe you or your situation?

1. I could handle a major unexpected expense
2. I am securing my financial future
3. Because of my money situation, I feel like I will never have the things I want in life*
4. I can enjoy life because of the way I’m managing my money
5. I am just getting by financially*
6. I am concerned that the money I have or will save won’t last*


How often does this statement apply to you?

7. Giving a gift for a wedding, birthday or other occasion would put a strain on my finances for the month*
8. I have money left over at the end of the month
9. I am behind with my finances*
10. My finances control my life*



The research question is clearly stated, can be answered by the data, and the context of the problem clearly explained.



## Review of the variables

The PUF includes financial attitude that were not included as part of the analysis but provide a interesting window into the relationship of financial attitudes to consumer financial wellbeing. 

```{r}

result_df <- getResultsCFPBFile()
result_df <- result_df %>% filter( cfpb_score > 0)

```




```{r}


# life_satisfied
ggplot(result_df, aes(x=life_satisfied)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "life_satisfied")

ggplot(data = result_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=life_satisfied, color=life_satisfied)) +
  labs (title = "life_satisfied")


ggplot(result_df,
    aes(cfpb_score, fill = life_satisfied, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by life_satisfied)")

```




```{r}

# life_optimistic
ggplot(result_df, aes(x=life_optimistic)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "life_optimistic")

ggplot(data = result_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=life_optimistic, color=life_optimistic)) +
  labs (title = "life_optimistic")


ggplot(result_df,
    aes(cfpb_score, fill = life_optimistic, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by life_optimistic)")
```



```{r}
# material_admire_wealth
ggplot(result_df, aes(x=material_admire_wealth)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "material_admire_wealth")

ggplot(data = result_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=material_admire_wealth, color=material_admire_wealth)) +
  labs (title = "material_admire_wealth")


ggplot(result_df,
    aes(cfpb_score, fill = material_admire_wealth, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by material_admire_wealth)")
```



```{r}

# material_define_worth
ggplot(result_df, aes(x=material_define_worth)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "material_define_worth")

ggplot(data = result_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=material_define_worth, color=material_define_worth)) +
  labs (title = "material_define_worth")


ggplot(result_df,
    aes(cfpb_score, fill = material_define_worth, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by material_define_worth)")

```



```{r}

# material_impress
ggplot(result_df, aes(x=material_impress)) +
  geom_bar(color="white", fill="black", alpha = 0.6, size = 0.1) +
  coord_flip() +
  labs(title = "material_impress")

ggplot(data = result_df) +
  geom_boxplot(mapping = aes(x=cfpb_score, y=material_impress, color=material_impress)) +
  labs (title = "material_impress")


ggplot(result_df,
    aes(cfpb_score, fill = material_impress, y = stat(density))) +
    geom_histogram(binwidth = 1, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(cfpb_score)), linetype = "dashed", size = 0.2) +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "CFPB Score (by material_impress)")

```








