---
title: DATA 606 Data Project Proposal
author: 'David Simbandumwe'
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("VennDiagram")
library(tidyverse)
library(readtext)
library(lubridate)

```




### Data Preparation

```{r}
# load data

crdt_df <- read_csv( file = "./CRDT_Data_CRDT.csv")
rt_df <- read_csv( file = "./rt.csv")
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Is race a determination factor in the likelihood of contracting COVID, being hospitalized or dying from COVID related medical complications. 






### Cases 

**What are the cases, and how many are there?**

There are 5320 unique cases identified by state and the specific date

### Data collection 

**Describe the method of data collection.**

The data was collected as part of the The COVID Racial Data Tracker project. The COVID Racial Data Tracker is a collaboration between the COVID Tracking Project and the Boston University Center for Antiracist Research. Together, we’re gathering the most complete and up-to-date race and ethnicity data on COVID-19 in the United States.

We’ve asked every state to report complete race and ethnicity data. Our Racial Data Dashboard has the latest.


### Type of study 

**What type of study is this (observational/experiment)?**

This is an observational study based on actual data reported by the 51 states / territorie.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The data was collected as part of the Covid Tracking Project https://covidtracking.com/data. For this project the data was downloaded as csv file for processing. 


Atlantic Monthly Group LLC (“The Atlantic”) 2021. 2021-03 [*The COVID Tracking Project at The Atlantic*](https://assets.ctfassets.net/o2ll9t4ee8tq/5P2QPF674OD1TGHUrVABx3/85a456045efa0d70ad42d9bf14c3beb0/CDC_Report_CTP.pdf) 

https://covidtracking.com/data

### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is the number of cases, deaths and hospitalizations


### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**
The independent variable is the race of the observations and it is qualitative / categorical


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

summary(crdt_df)

death_df <- crdt_df %>% 
  filter(Date > 20200101) %>%
  replace_na(list(Deaths_Total=0, Deaths_White=0, Deaths_Black=0, Deaths_Latinx=0, Deaths_Multiracial=0  )) %>%
  group_by(Date) %>% 
  mutate(
    date = ymd(Date),
    death_total = sum(Deaths_Total, na_rm=FALSE),
    death_white_total = sum(Deaths_White, na_rm=FALSE),
    death_black_total = sum(Deaths_Black, na_rm=FALSE),
    death_latinx_total = sum(Deaths_Latinx, na_rm=FALSE),
    death_multi_total = sum(Deaths_Multiracial, na_rm=FALSE)
  ) %>%
  select (date, death_total, death_white_total, death_black_total,death_latinx_total,death_multi_total) %>%
  distinct() 

summary(death_df)


ggplot(data = death_df) + 
  geom_line(mapping = aes(x=date, y=death_total, color="death_total")) +
  geom_line(mapping = aes(x=date, y=death_white_total, color="death white")) + 
  geom_line(mapping = aes(x=date, y=death_black_total, color="death black")) +
  geom_line(mapping = aes(x=date, y=death_latinx_total, color="death latinx")) +
  geom_line(mapping = aes(x=date, y=death_multi_total, color="death multi"))
  

  
```





```{r}

summary(rt_df)


rt_ny_df <- rt_df %>%
  filter(region == "NY") 
  
death_ny_df <- crdt_df %>% 
  filter(Date > 20200101, State=="NY") %>%
   mutate(
    date = ymd(Date)
   ) %>%
  replace_na(list(Deaths_Total=0, Deaths_White=0, Deaths_Black=0, Deaths_Latinx=0, Deaths_Multiracial=0  ))


ggplot(data = death_ny_df) + 
  geom_line(mapping = aes(x=date, y=Deaths_Total, color="death_total")) +
  geom_line(mapping = aes(x=date, y=Deaths_White, color="death white")) + 
  geom_line(mapping = aes(x=date, y=Deaths_Black, color="death black")) +
  geom_line(mapping = aes(x=date, y=Deaths_Latinx, color="death latinx")) +
  geom_line(mapping = aes(x=date, y=Deaths_Multiracial, color="death multi"))

ggplot(data = rt_ny_df) + 
  geom_line(mapping = aes(x=date, y=mean))

  




  

```


