---
title: "Lab 1: Intro to R"
author: "David Simbandumwe"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(dplyr)

data('arbuthnot', package='openintro')

```

### Exercise 1

```{r view-girls-counts}
arbuthnot$girls
```


### Exercise 2

The number of girls baptized over the years trends to rise consistenlty with the exception of 20 years between 1640 and 1660 where we observe a drop in baptizms. 

```{r trend-girls}
ggplot(data = arbuthnot, aes(x = year, y = girls)) + 
  geom_point()
```


### Exercise 3

Over the time period represented by the data there are consistently more boys than girls born in a single year. The percentage of boys born is a single year varies between just over 50% to over 53%.

```{r plot-prop-boys-arbuthnot}

arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls)

arbuthnot <- arbuthnot %>%
  mutate(boy_ratio = boys / total)

ggplot(data = arbuthnot, aes(x = year, y = boy_ratio)) + 
  geom_point()

```


### Exercise 4

The data sets includes the years 1940 to 2002 inclusively. The dimensions of the dataset are 3 columns with 63 rows of data. 

```{r dim-present}
data('present', package='openintro')
present

present %>% summarize(min = min(year), max = max(year))

```


### Exercise 5

The data sets includes the years 1620 to 1710 inclusively. The dimensions of the original data set are 3 columns with 82 rows of data. 

```{r count-compare}

arbuthnot

arbuthnot %>% summarize(min = min(year), max = max(year))


```


### Exercise 6

The authors observations do hold up there were more boys born in over the time period represented in the present data set. 

```{r plot-prop-boys-present}

present <- present %>%
  mutate(total = boys + girls)

present <- present %>%
  mutate(boy_ratio = boys / total)

ggplot(data = present, aes(x = year, y = boy_ratio)) + 
  geom_point()

```


### Exercise 7

In the data set the most births were recorded in 1961

```{r find-max-total}

present %>%
  arrange(desc(total))



```

